Scheduling Musical Events in Max/MSP with Scheme for Max

### 2021-06-24 paper-1/paper-1.tex
saving

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. 
The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C. 
The interpreter is instantiated
in the \textbf{s4m} object's constructor (\textbf{s4m\_new} as per the Max naming conventions), and
a reference to it is stored in the \textbf{s4m} object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated \textbf{s4m} object itself. Thus
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the \textbf{s4m} Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per \textbf{s4m} object instance, and
as s7 is thread safe, many \textbf{s4m} objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
thread and a low-priority "UI" thread for event/message flow. Each \textbf{s4m} object 
runs in only one of these, chosen via an optional constructor argument, and incoming messages
from the opposite thread are deferred or promoted accordingly.)

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. The
size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are designed to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

>>>>

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. 
The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C. 
The interpreter is instantiated
in the \textbf{s4m} object's constructor (\textbf{s4m\_new} as per the Max naming conventions), and
a reference to it is stored in the \textbf{s4m} object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated \textbf{s4m} object itself. Thus
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the \textbf{s4m} Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per \textbf{s4m} object instance, and
as s7 is thread safe, many \textbf{s4m} objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
thread and a low-priority ``UI" thread for event/message flow. Each \textbf{s4m} object 
runs in only one of these, chosen via an optional constructor argument, and incoming messages
from the opposite thread are deferred or promoted accordingly.)

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100 ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8 ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. 
% TODO could add citation here
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are designed to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.
